Drawing Trees and Animating Tree Changes

### 2010-05-14 paper/paper.tex
Submitted.

\begin{abstract} Dynamic trees appear in many areas of computer science,
including, but not limited to, Lisp code, data structures, tree automata, and
computer algorithms.  It would be useful to be able to easily draw trees and
animate them changing over time.  We begin by articulating a design for tree
visualization and animation that is more general than previous methodologies.
Our design is enabled by the observation that, for trees which are small enough
to be displayed on a screen, computers are now fast enough to eliminate many
worries about efficiency.  This observation allows us to make a library which
emphasizes simplicity of use over efficiency of execution.  We show how to use  our
library using several visualizations.\end{abstract}

\section{Introduction}
Trees, and especially trees which change over time, are fundamental objects in
computer science.  They occur in many disparate contexts, and while the uses of
these trees may be different, there exist commonalities which we can use to
build a tree visualization system which might be useful in many different
contexts.  We will articulate the design principles of a highly usable
tree visualization system, present one such system that we have built in Java
and made freely available, and then show our system's utility with
several sample visualizations.

There are multiple reasons to visualize a system, and different reasons have
different requirements.  A visualization designed for experts may be very
useful for that target group, but leave neophytes confused.  Therefore, before
we discuss the particulars of our visualization, we must explain our reasons
for wanting such a system in the first place, and the expected audience.  We
want to visualize dynamic trees for demonstration, education, and beauty, and
we would like our visualizations to be of benefit to computer science
undergraduates.

@@

One reason for visualization is for demonstration.  Trees (and other systems)
are visualized in an effort to show how these systems work.  Because this is
one of our goals, we will restrict ourselves to trees which are small enough to
be drawn in their entirety on a single computer screen.  Whenever we visualize a tree, our display system will show the entire tree, and it will be part of the job of our library to make the best use of the available screen real estate.

@@

Computing and math educators have a long history of building visualization
systems for demonstration.  The intuition in many case was that if students can view these
systems in action, then the students will understand these system.
Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
visualizations are useless\cite{hds}.  A good visualization seems intuitively useful, but
it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.  The one bright spot in this was in interactivity and construction.  The one activity that improved student understanding was not the viewing of a visualization, but instead its construction.

If students, instead of viewing a visualization, construct their own, then they
learn.  Unfortunately, visualizations are generally desired when a system is
sufficiently complex to be difficult to understand.  If a system is difficult
to understand, then it is difficult to program, and, even worse, graphics
programming is perceived as one of the trickiest programming tasks.  This means
that when we ask students to build a visualization, we are asking them to take
a system they do not understand, build it correctly, and then engage in an
extremely delicate programming task with their (hopefully correct, but likely
buggy) system.

For effective educational visualization, students must write their own
visualizations.  Therefore, whenever possible, we design our system to be as
simple, learnable, and robust as possible.  This will allow more students to
succeed in the creation of visualizations by attempting to make the graphics
aspect of the programming task as simple as possible.

@@

To achieve this goal of beauty, we will pay attention to the aesthetic aspects
of our visualization as well as maintaining correctness.  In particular, we
should ensure that transitions are smooth and not jarring, and that our
layouts are as aesthetically pleasing as possible.  This requirement is
difficult to quantify, but, cetus paribus, a beautiful visualization is always
better than an ugly one.

@@

Let us begin our design by formally defining the objects under consideration.
A \newword{tree} $T$ is an undirected connected acyclic graph with vertex set
$V$ and edge set $E \subseteq \{V \times V\}$.  A \newword{rooted tree} is a
tree where a particular vertex $r$ is designated the as the \newword{root} of
the tree, which allows us to define the \newword{depth} of a vertex $v$ as the
length of the shortest path between $v$ and $r$. Every edge connects two
vertices $(u,v)$, and we will assume, again without loss of generality, that the depth of $u$ is one
less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
$v$ and $v$ the \newword{child} of $u$.  The \newword{children} of $v$ are the
set of vertices which have $v$ as their parent.  A \newword{rooted ordered
tree} is a rooted tree with a partial-ordering on the vertex set $V$ such that
for all vertices $\{u,v, \ldots\}$ with the same parent, either $u < v$ or $v <
u$.

Without loss of generalization, we restrict our visualization system to rooted,
ordered trees.  Any tree can be converted into a rooted, ordered tree by
selecting a root and imposing a partial-ordering on the vertices.  In practice,
however, we found that every domain we considered had a naturally occurring root
and ordering, and that the imposition of an arbitrary root or an arbitrary
ordering was unnecessary.  From here forward, when we say tree, we will mean a
rooted ordered tree.

\section{Constraints of Tree Layout}
\label{sec:constraints}
To lay out a tree in the plane, we must map every vertex $v \in V$ to a
coordinate $(v_x, v_y)$.  When visualizing a tree we put the root at the top
(which, while counter-intuitive for biologists, is standard practice in
computer science), and every vertex $\{u,v \ldots\} \subset V$ with equal depth
should be laid out with the same $y$-coordinate $(u_y = v_y)$.  Furthermore, if
two nodes $u$ and $v$ have the same parent and $u < v$, then we should lay them
out with $u_x < v_x$, which will preserve in our layout the natural order of the tree.  Our last requirement is that if we have two nodes $w$ and $z$ with equal depth and children $U$ and $V$, and $w_x < z_x$, then $u_x < v_x \forall u\in U, v\in V$.  This will avoid crossing edges, which is a good thing because trees are planar graphs, and so drawing them with crossing edges is totally unnecessary.

@@

Tree animation has a rich body of work associated with it.  In particular, much
past work has focused on exactly what tree operations may be supported
efficiently\cite{Moen,tamassia} as well as on the force-directed placement
algorithm for tree layout\cite{fdp, Cohen}.  The entirety of this work has been
largely spent on designing systems in which trees can be laid out and animated
in an efficient manner.  Our work differs from theirs due to our focus on
ease-of-use, and that focus comes from the fact that the speed of a ``current''
computer has grown dramatically since the original papers were written.

@@

Either on every tree transformation or when requested by the visualization
programmer, we re-layout (off-screen) the entire tree according to our layout
algorithm and then we animate the transition from the old tree to the new
tree.  Or animation procedure is extremely basic, but the simplicity allows for simplicity of use.

Our animation procedure has six steps.  First, we lay out, off-screen, the new
tree.  Second, we zoom out (if required) so that the viewable area is the size
of the maximum of the new tree and old tree.  Third, we fade out any of the
vertices and edges which were in the old tree, but have been removed in the new
tree. Fourth, we animate the movement of all vertices which are present in both
trees from their old positions to their new positions.  Fifth, we fade in any
new nodes which are present only in the new tree.  Sixth, we zoom in (if
required) so that the new tree fills the viewing area.

Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.

>>>>

\begin{abstract} Dynamic trees appear in many areas of computer science,
including, but not limited to, Lisp code, data structures, tree automata, and
computer algorithms.  It would be useful to be able to easily draw trees and
animate them changing over time.  We articulate a design for tree
visualization and animation that is more general than previous methodologies.
Our design is enabled by the observation that, for trees which are small enough
to be displayed on a screen, computers are now fast enough to eliminate many
concerns about efficiency.  This observation allows us to make a library which
emphasizes simplicity of use over efficiency of execution.  We show how to use  our
library using several visualizations.\end{abstract}

\section{Introduction}
Trees, and especially trees which change over time, are fundamental objects in
computer science.  They occur in many disparate contexts, and while the uses of
these trees may be different, there exist commonalities which we exploit to
build a tree visualization system which might be useful in many different
contexts.  We will articulate the design principles of a highly usable
tree visualization system, present one such system that we have built in Java
and made freely available, and then show our system's utility with
several sample visualizations.

There are multiple reasons to visualize a system, and different reasons have
different requirements.  A visualization designed for experts may be very
useful for that target group, but leave neophytes confused.  Therefore, before
we discuss the particulars of our visualization, we must explain our reasons
for wanting such a system in the first place, as well as identify the expected
audience.  We want to visualize dynamic trees for demonstration, education, and
beauty, and we would like our system to be of benefit to computer
science undergraduates.

@@

One reason for visualization is for demonstration.  Trees and other systems
are visualized in an effort to show how these systems work.  Because this is
one of our goals, we will restrict ourselves to trees which are small enough to
be drawn in their entirety on a single computer screen.  Whenever we visualize a tree, our display system will show the entire tree, and it will be part of the job of our library to make the best use of the available screen real estate.

@@

Computing and math educators have a long history of building visualization
systems for demonstration.  The intuition was that if students could view these
systems in action, then the students would understand these systems.
Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
visualizations are useless\cite{hds}.  A good visualization seems intuitively helpful, but
it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.  The one bright spot in this was in interactivity and construction.  The activity that improved student understanding was not the viewing of a visualization, but instead its construction and interactive use.

If students, instead of viewing a visualization, construct their own, then they
learn.  Unfortunately, visualizations are generally desired when a system is
sufficiently complex to be difficult to understand.  If a system is difficult
to understand, then it is difficult to program, and, even worse, graphics
programming can be one of the trickiest programming tasks.  This means
that when we ask students to build a visualization, we are asking them to take
a system they do not understand, build it correctly, and then engage in an
extremely delicate programming task with their (hopefully correct, but likely
buggy) system.

For effective educational visualization, students must write their own
visualizations.  Therefore, whenever possible, we design our system to be as
simple, learnable, and robust as possible.  This will allow more students to
succeed in the creation of visualizations by making the graphics
aspect of the programming task as simple as possible.

@@

This requirement is difficult to quantify, but, cetus paribus, a beautiful
visualization is always better than an ugly one.  To achieve this goal of
beauty, we will pay attention to the aesthetic aspects of our visualization as
well as maintaining correctness.  In particular, we should ensure that
transitions are smooth and not jarring, and that our layouts are as
aesthetically pleasing as possible.

@@

Let us begin our design by formally defining the objects under consideration.
A \newword{tree} $T$ is an undirected connected acyclic graph with vertex set
$V$ and edge set $E \subseteq \{V \times V\}$.  A \newword{rooted tree} is a
tree where a particular vertex $r$ is designated the as the \newword{root} of
the tree, which allows us to define the \newword{depth} of a vertex $v$ as the
length of the shortest path between $v$ and $r$. Every edge connects two
vertices $(u,v)$, and we will assume, without loss of generality, that the depth of $u$ is one
less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
$v$ and $v$ the \newword{child} of $u$.  The \newword{children} of $v$ are the
set of vertices which have $v$ as their parent.  A \newword{rooted ordered
tree} is a rooted tree with a partial-ordering on the vertex set $V$ such that
for all vertices $\{u,v, \ldots\}$ with the same parent, either $u < v$ or $v <
u$.

We restrict our visualization system to rooted,
ordered trees.  Any tree can be converted into a rooted, ordered tree by
selecting a root and imposing a partial-ordering on the vertices.  In practice,
however, we found that every domain we considered had a naturally occurring root
and ordering, and that the imposition of an arbitrary root or an arbitrary
ordering was unnecessary.  From here forward, when we say tree, we will mean a
rooted, ordered tree.

\section{Constraints of Tree Layout}
\label{sec:constraints}
To lay out a tree in the plane, we must map every vertex $v \in V$ to a
coordinate $(v_x, v_y)$.  When visualizing a tree we put the root at the top
(which, while counter-intuitive for biologists, is standard practice in
computer science), and every vertex $\{u,v \ldots\} \subset V$ with equal depth
should be laid out with the same $y$-coordinate $(u_y = v_y)$.  Furthermore, if
two nodes $u$ and $v$ have the same parent and $u < v$, then we should lay them
out with $u_x < v_x$, which will preserve in our layout the natural order of the tree.  Our last requirement is that if we have two nodes $w$ and $z$ with equal depth and children $U$ and $V$, and $w_x < z_x$, then $u_x < v_x \forall u\in U, v\in V$.  This will avoid crossed edges, which is a good thing because trees are planar graphs, and so drawing them with crossing edges is totally unnecessary.

@@

Tree animation has a rich body of work associated with it.  In particular, much
past work has focused on exactly what tree operations may be supported
efficiently\cite{Moen,tamassia} as well as on the force-directed placement
algorithm for tree layout\cite{fdp, Cohen}.  The entirety of this work has been
largely spent on designing systems in which trees can be laid out and animated
in an efficient manner.  Our work differs from theirs due to our focus on
ease-of-use, and that focus comes from the fact that the speed of commonly
available computers has grown dramatically since the original papers were
written.

@@

Either on every tree transformation or when requested by the visualization
programmer, we re-layout (off-screen) the entire tree according to our layout
algorithm and then we animate the transition from the old tree to the new tree.
Or animation procedure is extremely simple, but that simplicity allows for
simplicity of use.

Our animation procedure has six steps.  First, we lay out, off-screen, the new
tree.  Second, we zoom out (if required) so that the viewable area is the size
of the maximum required for the new tree and old tree.  Third, we fade out any of the
vertices and edges which were in the old tree, but have been removed in the new
tree. Fourth, we animate the movement of all vertices which are present in both
trees from their old positions to their new positions.  Fifth, we fade in any
new nodes which are present only in the new tree.  Sixth, we zoom in (if
required) so that the new tree fills the viewing area.

Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default.  This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.
